---
title: "HW3_yc4384_Cynthia"
author: "Yangyang Chen"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    keep_tex: true
header-includes:
  - \usepackage{booktabs}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(psych)
```
## Problem 1

### (a) Fit a prospective model to the data to study the relation between alcohol consumption, age, and disease (model age as a continuous variable taking values 25, 35, 45, 55, 65, and 75). Interpret the result.

Using logistics regression model to fit data from prospective study:

$$log(\frac{\pi}{1-\pi}) = \beta_0 +\beta_1X_{alc}+\beta_2X_{age}$$
```{r}
# load data
age = seq(from = 25, to = 75, by = 10) |> 
  rep(2)
case = c(1, 4, 25, 42, 19, 5, 0, 5, 21, 34, 36, 8)
control = c(9, 26, 29, 27, 18, 0, 106, 164, 138, 139, 88, 31)
alc = c(rep(1,6), rep(0, 6))
resp = cbind(case, control)

# Model fitting using logit link
glm_logit=glm(resp ~ alc + age, family=binomial(link='logit'))
summary(glm_logit)
```
Hence, the logistics regression model is:
$$log(\frac{\pi}{1-\pi}) = -5.02 + 1.78X_{alc} + 0.06*X_{age}$$

Interpretation:

* The model suggests a significant relationship between esophageal cancer, daily alcohol consumption adjusted and age. 

* $\beta_{1}$: the log odds ratio of having the disease among heavy drinkers is $1.78$ times the odds odds ratio of non-heavy drinkers, keeping age fixed.

* $exp(\beta_{1})$: odds ratio for the association between disease and alcohol consumption, holding age constant.

* $\beta_{2}$: the log odds ratio of having the disease will increase by $0.06$ for every unit increment in age, keeping alcohol consumption fixed.

* $exp(\beta_{2})$: the odds ratio for the association between disease and age, holding alcohol consumption constant.

* This model appears to fit the data well, as indicated by the significant coefficients and the reduction in deviance from the null model to the fitted model.

(b)
```{r}
age = c(1:6) |> 
  factor()
ind = dummy.code(age)
grp1 = rep(ind[,1],2)
grp2 = rep(ind[,2],2)
grp3 = rep(ind[,3],2)
grp4 = rep(ind[,4],2)
grp5 = rep(ind[,5],2)
grp6 = rep(ind[,6],2)

M_0 = glm(resp ~ grp1 + grp2 + grp3 + grp4 + grp5 + grp6, family = binomial(link = 'logit'))
summary(M_0)
dev_m0 = residuals(M_0, type = "deviance")^2 |> sum()

M_1 = glm(resp ~ alc + grp1 + grp2 + grp3 + grp4 + grp5 + grp6, family = binomial(link = 'logit'))
summary(M_1)

# Define the data
Model <- c("$M_0$", "$M_1$")
Linear_predictor <- c("$\\alpha_j$", "$\\alpha_j + \\beta X$")
Deviance <- c(90.56, 11.04)
df <- c(6, 5)

# Create a data frame
outp <- data.frame(Model, Linear_predictor, Deviance, df, stringsAsFactors = FALSE)

# Set column names
colnames(outp) <- c("Model", "Linear Predictor", "Deviance", "df")

# Print the data frame using knitr::kable() with format = "latex" and escape = FALSE

knitr::kable(outp, format = "latex", escape = FALSE, align = "c", booktabs = TRUE)
```
Hypothesis Testing:

$$H_0: \beta_{alc}=0 \\ \ H_1:\beta_{alc} \ne 0 $$
$$stat = 79.52 \sim \chi^2_{1}$$
$$p-value <0.0001$$

Hence, we reject the null hypothesis and we have sufficient evidence to conlcude that there's an significant association between disease and alcohol consumption.

## Problem 2

### (a) Fit a logistic regression model to study the relation between germination rates and different types of seed and root extract. Interpret the result.

Let $Y_i$ denote the number of seeds germinates among $m_i$ seeds with the $ith$ covariate pattern. The logistic regression model:

$$log(\frac{\pi}{1-\pi}) = \beta_0 + \beta_1X_{seed}+\beta_2X_{root};$$

```{r}
# input data
x=c(rep(1,10),rep(0,11)) # roots category
z=c(rep(1,5), rep(0,6), rep(1,5), rep(0,5)) # seeds category
y=c(10,23,23,26,17,8,10,8,23,0,5,53,55,32,46,10,3,22,15,32,3) # survive=1
m=c(39,62,81,51,39,16,30,28,45,4,6,74,72,51,79,13,12,41,30,51,7)
data=data.frame(x,y,m)
plot(x,m)
plot(y,m)
summary(m-y) ## m >= y

# fit binomial (logistic) without dispersion
none.disp=glm(cbind(y,m-y)~x+z, family=binomial(link='logit'))
summary(none.disp)
G.stat=sum(residuals(none.disp,type='pearson')^2) # pearson chisq 
G.stat
```
The binomial logistic without dispersion was fitted with the following results:

$$log(\frac{\pi}{1-\pi}) = 0.3884 - 1.0668*X_{root}+0.2391*X_{seed}$$

Interpretation:
 
 $$\hat{b_0} = 0.3884,\ standard \ error(b_0) = 0.1410;$$
 
* The log odds ratio of a O.aegyptiaca 73 seed grown in cucumber extract for germinating is 0.38.

$$\hat{b_1} = 0.2391,\ standard \ error(b_1) = 0.1538;$$

* The estimated log odds ratio for comparing O.aegyptiaca 75 seeds and O.aegyptiaca 73 seeds, holding root extract fixed is 0.2391.

$$\hat{b_2} = -1.0668,\ standard \ error(b_2) = 0.1442;$$

* The estimated odds ratio for comparing bean and cucumber extract amongst O.aegyptiaca 73 seeds, holding seed species fixed is -1.0668.


$$Pearson-\chi^2 statistic: X^2 = \sum X_i^2 = 41.226 \ and\ Deviance \ D = \sum d_i^2 = 40.328.$$

```{r}
# goodness of fit
pval=1-pchisq(none.disp$deviance,21-3) 
pval # bad fit, reject the fitting
```
* Comparing $X^2$ and $D$ with $\chi^2(18)$, we concluded that the model appears to fit bad.


### (b) Is there over dispersion? If so, what is the estimate of dispersion parameter? Update your model and reinterpret the result.

Estimating the dispersion parameter by following two methods:

First,$$\hat{\phi} = G_0/(n-p),$$
where $$G_0 = \sum_{i=1}^{n}{\frac{(y_i-m_i \hat{\pi_i})^2}{m_i\hat{\pi_i}(1-\hat{\pi_i})\phi}} \sim \chi^2(n-p)\ $$
is the generalized Pearson $\chi^2$ from the original model fitting without over-dispersion.

Second, $$\hat{\phi} = \frac{D_0}{n-p}$$
```{r}
# calc dispersion para in 2 methods
# the first method
phi=G.stat/(21-3)
phi

# the second method
tilde.phi=none.disp$deviance/none.disp$df.residual
tilde.phi # similar to the one estimated from pearson chisq 

# test over-dispersion (half normal plot)
res=residuals(none.disp,type='pearson')
plot(qnorm((21+1:21+0.5)/(2*21+1.125)),sort(abs(res)),xlab='Expected Half-Normal Order Stats',ylab='Ordered Abs Pearson Residuals')
abline(a=0,b=1)
abline(a=0,b=sqrt(phi),lty=2, col = 'red')
```

* Therefore, there exists over-dispersion in our model and the estimate of dispersion parameter: $\hat{\phi} = 2.1697$. 

* Half-normal plot using residuals
from this model shows evidence of over-dispersion.

* Next, we updated regression model.

```{r}
# fit model with constant over-dispersion
summary(none.disp,dispersion=phi)
```
The binomial logistic with dispersion was fitted with the following results:
$$b_0 = 0.3884,\ standard \ error(b_1) = 0.2071;$$
$$b_1 = -1.0668,\ standard \ error(b_1) = 0.2119;$$
$$b_2 = 0.2391,\ standard \ error(b_2) = 0.2259;$$
$$Pearson-\chi^2 statistic: X^2 = \sum X_i^2 = 41.226 \ and\ Deviance \ D = \sum d_i^2 = 40.328.$$

### (c) What is a plausible cause of the over dispersion?

Since different groups may have different germination rate which follow the same distribution, the response variables should follows a beta-binomial distribution rather than binomial distribution.

